---
title: Analysis
description: Here we provide a detailed analysis using more sophisticated statistics techniques.
toc: true
draft: false
---
# The question we want to address 

- The primary focus of this research is to explore the relationship between an individual's total income and their rental expenditures.
  - The analysis aims to determine how rental spending adjusts in response to changes in income levels.
  - Both linear and nonlinear models are employed to elucidate the connection between these variables.
  - Additionally, the study considers racial differences, examining how the coefficients vary among different racial groups.
  - A time series regression model is utilized to account for temporal variations in this relationship.
  
# Modeling Approaches Explored

- Simple Linear Regression Model (Utilizing the 2022 Dataset)
  - Initial fitting of this model was conducted to determine the potential worth of exploring the relationship between the predictor and response. The findings indicated that further investigation was merited.

- Multiple Linear Regression Model (Utilizing the 2022 Dataset)
  - This extended model incorporated additional confounding variables, building upon the simple linear regression.
  - A multicollinearity assessment was performed to circumvent overfitting issues.
  - The selection of control variables was affirmed.
  - The residual plot indicated the presence of heteroscedasticity.

- Log-Log Regression Model (Utilizing the 2022 Dataset)
  - Both the predictor and response variables underwent log transformation to address the problem in the Multiple Linear Regression Model.
  - Outliers were identified and removed from the analysis.
  - All variables demonstrated statistical significance.
  - The residual plot suggested an acceptable fit.
  
- Group_by **race**, log_log regression model (using 2022 dataset) -- Detail in class blog 5

- Group_by **race**, log_log regression model (using 2000-2022 dataset) -- Detail in class blog 6

- Time series model (using 2000-2022 dataset) -- Detail in class blog 7



##  Modeling and Inference


## Rough explorationn to the correlation 
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyr)
library(broom)
library(dplyr)
library(here)
library(gt)
library(purrr)
library(readr)
library(ggplot2)
library(knitr)
data_path <- here("dataset", "data_twenty_clean.csv")
data_path1 <- here("dataset", "latest_clean.csv")
data_twenty <- read_csv(data_path,show_col_types = FALSE) 
data_2022 <- read_csv(data_path1,show_col_types = FALSE) 
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data_2022, aes(x = INCTOT, y = RENTGRS)) +
  geom_point(color = 'black', alpha = 0.4, size = 1.5) +  
  geom_smooth(method = "lm", color = "darkblue", size = 1.2) +  
  scale_x_continuous(
    limits = c(0, quantile(data_2022$INCTOT, 0.95)),
    labels = scales::comma,  
    name = "Total Income"  
  ) +
  scale_y_continuous(
    limits = c(0, quantile(data_2022$RENTGRS, 0.95)),
    labels = scales::comma,
    name = "Gross Rent"  
  ) +
  labs(title = "Relationship Between Total Income and Gross Rent of 2022 dataset",  
       subtitle = "Linear model fit shown in blue",
       caption = "Data source: IPUMS 2022") +
  theme_minimal() +  
  theme(
    text = element_text(size = 12), 
    plot.title = element_text(face = "bold"), 
    axis.title = element_text(face = "bold")  
  )
```

- There seems to be a large concentration of points towards the lower end of the income scale with lower rents, which suggests that a significant portion of the dataset includes individuals with lower income and rent levels.

- For the assumptions of linear model, the plot only provide rough sense of what we could consider
  - The trend line suggests a possible linear relationship, but this needs further statistical testing.
  - The spread of rent appears to increase with income, hinting at heteroscedasticity rather than the required constant variance across levels of income.
  - Normality of Residuals: Cannot be assessed from the plot; but as the data_page indicate, we make log transformation 
  -  There are visible potential outliers, especially at higher income levels, which could affect the model's accuracy. We also found this problem in the data_page, therefore we exclude the outlier which defined as the range outside of the three standard deviations. 








  


------------------------------------------------------------------------

## Rubric: On this page

You will

-   Introduce what motivates your Data Analysis (DA)
    -   Which variables and relationships are you most interested in?
    -   What questions are you interested in answering?
    -   Provide context for the rest of the page. This will include figures/tables that illustrate aspects of the data of your question.
-   Modeling and Inference
    -   The page will include some kind of formal statistical model. This could be a linear regression, logistic regression, or another modeling framework.
    -   Explain the ideas and techniques you used to choose the predictors for your model. (Think about including interaction terms and other transformations of your variables.)
    -   Describe the results of your modelling and make sure to give a sense of the uncertainty in your estimates and conclusions.
-   Explain the flaws and limitations of your analysis
    -   Are there some assumptions that you needed to make that might not hold? Is there other data that would help to answer your questions?
-   Clarity Figures
    -   Are your figures/tables/results easy to read, informative, without problems like overplotting, hard-to-read labels, etc?
    -   Each figure should provide a key insight. Too many figures or other data summaries can detract from this. (While not a hard limit, around 5 total figures is probably a good target.)
    -   Default `lm` output and plots are typically not acceptable.
-   Clarity of Explanations
    -   How well do you explain each figure/result?
    -   Do you provide interpretations that suggest further analysis or explanations for observed phenomenon?
-   Organization and cleanliness.
    -   Make sure to remove excessive warnings, hide most or all code, organize with sections or multiple pages, use bullets, etc.
    -   This page should be self-contained, i.e. provide a description of the relevant data.
