---
title: Data
description: We describe the sources of our data and the cleaning process.
toc: true
draft: false
---
## Describe where/how to find data.
Our dataset comes from [IPUMS USA](https://usa.ipums.org/usa/index.shtml). 

> Reference:
Steven Ruggles, Sarah Flood, Matthew Sobek, Daniel Backman, Annie Chen, Grace Cooper, Stephanie Richards, Renae Rodgers, and Megan Schouweiler. IPUMS USA: Version 15.0 [dataset]. Minneapolis, MN: IPUMS, 2024.
https://doi.org/10.18128/D010.V15.0

The original data of this website came from censuses as well as the American Community Surveys (ACS) that began in 2000 and are still going on today. We chose our dataset from this website IPUMS because this website contains datasets that are large, valid and complex, containing detailed microdata, which aims to facilitate transformative research, education, and policy-making by aggregating data across times and places. It makes the datasets on this website very appropriate for our research to streamline our analysis of changes, data merging, and comparative research.
  
  
## Describe the different data files used and what each variable means.

### Table contains all the variables we have

We have highlighted all key dependent variables (RENTGRS) and independent variables (INCTOT) in bold.

| Variable | Description                           |
|----------|---------------------------------------|
| **RENTGRS**  | Monthly gross rent                |
| KITCHEN  | Kitchen or cooking facilities         |
| ROOMS    | Number of rooms                       |
| NFAMS    | Number of families in household       |
| AGE      | Age                                   |
| MARST    | Marital status                        |
| RACE     | Race                                  |
| EMPSTAT  | Employment status                     |
| **INCTOT**   | Total personal income             |
| FTOTINC  | Total family income                   |
| REGION   | Geographical part of America          |


## Describe any cleaning you had to do for your data.
- Here is the link to our cleaned dataset [cleaned dataset.](./dataset//latest.rds)

- The code for cleaning and recoding the datasets, please see class blog 2 and 3. 

- For our dependent variable—rent—we exclude records indicating a rent of 0. This exclusion is based on the understanding that a 0 rent suggests the individual owns their home or lives in a situation not pertinent to our target research population, which is individuals who rent their housing.

- Exclude data entries characterized by specific codes indicative of missing or unknown values, as well as those representing values deemed inappropriate for analytical purposes. The following table shows the variables we processed:
* Given the vast size of the original dataset, which includes millions of records, we opted to remove these entries directly. This decision does not compromise our ability to conduct a thorough analysis, as the remaining dataset is still large and clean enough for reliable and robust research conclusions.

| Variable     | Description                           | Notes                                                                                                   |
|--------------|---------------------------------------|---------------------------------------------------------------------------------------------------------|
| **INCTOT**   | Total personal income                 | Specific codes detailed below:                                                                          |
|              |                                       | 0000001 = $1 or break even (2000, 2005-onward ACS and PRCS)                                             |
|              |                                       | 9999999 = N/A                                                                                           |
|              |                                       | 9999998 = Unknown                                                                                       |
| **FTOTINC**  | Total family income                   | Specific codes detailed below:                                                                          |
|              |                                       | -000001 = Net loss (1950)                                                                               |
|              |                                       | 9999998 = Not ascertained (1950)                                                                        |
|              |                                       | 9999999 = N/A                                                                                           |
| **KITCHEN**  | Kitchen or cooking facilities         | 0 = N/A                                                                                                 |
| **ROOMS**    | Number of rooms                       | 00 = N/A                                                                                                |
| **EMPSTATA** | Employment status                     | Specific codes detailed below:                                                                          |
|              |                                       | 0 = Missing value                                                                                       |
|              |                                       | 9 = Unknown/Illegible                                                                                   |
| **STATE**    | Geographic state identifier           | Specific codes detailed below:                                                                          |
|              |                                       | 91 = Overseas Military/Military Installations                                                           |
|              |                                       | 92 = PUMA boundaries cross state lines - Metro sample                                                   |
|              |                                       | 97 = State not identified                                                                               |
|              |                                       | 99 = Not identified                                                                                     |


### Recode some categorical data that is showed in numerical value in our raw data.

- Recode certain variables. We modified certain variables in the dataset that are encoded with specific numerical codes, which reflect categories or qualitative information rather than quantitative values.
The following table shows each variable that had been recoded;
The detailed process of recoding is in Class blog3.

| Variable     | Description                               | Notes                                                                 |
|--------------|-------------------------------------------|-----------------------------------------------------------------------|
| **INCTOT**   | Total personal income                     | Specific codes detailed below:                                        |
|              |                                           | -009995 = -$9,900 (1980)                                              |
|              |                                           | -000001 = Net loss (1950)                                             |
|              |                                           | FTOTINC-000001 = Net loss (1950)                                      |
| **KITCHEN**  | Kitchen facilities                        | 0 = N/A                                                               |
|              |                                           | 1 = No                                                                |
|              |                                           | 2 = No, or shared use                                                 |
|              |                                           | 3 = Yes, shared use                                                   |
|              |                                           | 4 = Yes (shared or exclusive use)                                     |
|              |                                           | 5 = Yes, exclusive use                                                |
| **MARST**    | Marital status                            | 1 = Married, spouse present                                           |
|              |                                           | 2 = Married, spouse absent                                            |
|              |                                           | 3 = Separated                                                         |
|              |                                           | 4 = Divorced                                                          |
|              |                                           | 5 = Widowed                                                           |
|              |                                           | 6 = Never married/single                                              |
|              |                                           | 9 = Blank, missing                                                    |
| **RACE**     | Race                                      | 1 = White                                                             |
|              |                                           | 2 = Black/African American                                            |
|              |                                           | 3 = American Indian or Alaska Native                                  |
|              |                                           | 4 = Chinese                                                           |
|              |                                           | 5 = Japanese                                                          |
|              |                                           | 6 = Other Asian or Pacific Islander                                   |
|              |                                           | 7 = Other race, nec                                                   |
|              |                                           | 8 = Two major races                                                   |
|              |                                           | 9 = Three or more major races                                         |
| **EMPSTAT**  | Employment status                         | 1 = Employed                                                          |
|              |                                           | 2 = Unemployed                                                        |
|              |                                           | 3 = Not in the labor force                                            |
| **REGION**   | Geographic region and division            | Specific codes detailed below:                                        |
|              |                                           | 11 = New England Division (Northeast)                                 |
|              |                                           | 12 = Middle Atlantic Division (Northeast)                             |
|              |                                           | 13 = Mixed Northeast Divisions (Northeast)                                        |
|              |                                           | 21 = East North Central Division (Midwest)                            |
|              |                                           | 22 = West North Central Division (Midwest)                            |
|              |                                           | 23 = Mixed Midwestern Divisions (Midwest)                                       |
|              |                                           | 31 = South Atlantic Division (South)                                  |
|              |                                           | 32 = East South Central Division (South)                              |
|              |                                           | 33 = West South Central Division (South)                              |
|              |                                           | 34 = Mixed Southern Divisions (South)                                        |
|              |                                           | 41 = Mountain Division (West)                                         |
|              |                                           | 42 = Pacific Division (West)                                          |
|              |                                           | 43 = Mixed Western Divisions (West)                                         |



#### Merged data
We consolidated datasets spanning 22 years, from 1990 to 2022, sourced from the same repositories. The approach to cleaning and processing these datasets was aligned with the methodology applied to the 2022 dataset, as previously detailed.
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(dplyr)
data_path <- here("dataset-ignore", "twenty_yr_data.csv")
library(readr)
twentyyr_data <- read_csv(data_path,show_col_types = FALSE)
library(dplyr)
set.seed(123) # Setting a seed for reproducibility
twentyyr_data_clean <- sample_frac(twentyyr_data, 0.01)
twentyyr_data_clean <- twentyyr_data_clean %>%
  filter(RENTGRS != 0,  # Exclude entries with zero rent
         !is.na(RENTGRS),  # Exclude NA values for rent
         EMPSTAT %in% c("1", "2", "3"),  # Include only valid employment statuses
         KITCHEN != "0",  # Exclude entries without a kitchen
         ROOMS != "00",  # Exclude entries with invalid room counts
         !INCTOT %in% c(1, 9999999, 9999998, -1),  # Exclude specific INCTOT values
         !FTOTINC %in% c(9999998, 9999999),  # Exclude specific FTOTINC values
         POVERTY != 0,  # Exclude invalid poverty codes
         !REGION %in% c(91, 92, 97, 99))  # Exclude specific region/state codes
twentyyr_data_clean$REGION_CLASSIFIED <- case_when(
  twentyyr_data_clean$REGION %in% c(11, 12, 13) ~ "NORTHEAST",
  twentyyr_data_clean$REGION %in% c(21, 22, 23) ~ "MIDWEST",
  twentyyr_data_clean$REGION %in% c(31, 32, 33, 34) ~ "SOUTH",
  twentyyr_data_clean$REGION %in% c(41, 42, 43) ~ "WEST",
  TRUE ~ "OTHER"
)

#KITCHEN
twentyyr_data_clean <- twentyyr_data_clean %>%
  mutate(KITCHEN = case_when(
    KITCHEN == "0" ~ "N/A",
    KITCHEN == "1" ~ "No",
    KITCHEN == "2" ~ "No, or shared use",
    KITCHEN == "3" ~ "Yes, shared use",
    KITCHEN == "4" ~ "Yes (shared or exclusive use)",
    KITCHEN == "5" ~ "Yes, exclusive use",
    TRUE ~ as.character(KITCHEN)
  ))
twentyyr_data_clean$KITCHEN <- as.factor(twentyyr_data_clean$KITCHEN)

#MARST
twentyyr_data_clean <- twentyyr_data_clean %>%
  mutate(MARST = case_when(
    MARST == "1" ~ "Married, spouse present",
    MARST == "2" ~ "Married, spouse absent",
    MARST == "3" ~ "Separated",
    MARST == "4" ~ "Divorced",
    MARST == "5" ~ "Widowed",
    MARST == "6" ~ "Never married/single",
    MARST == "9" ~ "Blank, missing",
    TRUE ~ as.character(MARST)
  ))
twentyyr_data_clean$MARST <- as.factor(twentyyr_data_clean$MARST)

#RACE
twentyyr_data_clean <- twentyyr_data_clean %>%
  mutate(RACE = case_when(
    RACE == "1" ~ "White",
    RACE == "2" ~ "Black/African American",
    RACE == "3" ~ "American Indian or Alaska Native",
    RACE == "4" ~ "Chinese",
    RACE == "5" ~ "Japanese",
    RACE == "6" ~ "Other Asian or Pacific Islander",
    RACE == "7" ~ "Other race, nec",
    RACE == "8" ~ "Two major races",
    RACE == "9" ~ "Three or more major races",
    TRUE ~ as.character(RACE) 
  ))
twentyyr_data_clean$RACE <- as.factor(twentyyr_data_clean$RACE)

#EMPSTAT
twentyyr_data_clean <- twentyyr_data_clean %>%
  mutate(EMPSTAT = case_when(
    EMPSTAT == "1" ~ "Employed",
    EMPSTAT == "2" ~ "Unemployed",
    EMPSTAT == "3" ~ "Not in labour force",
    TRUE ~ as.character(EMPSTAT) 
  ))

twentyyr_data_clean$EMPSTAT <- as.factor(twentyyr_data_clean$EMPSTAT)

write_csv(twentyyr_data_clean, file = here("dataset", "twentyyear_rent.csv"))
saveRDS(twentyyr_data_clean, file = here("dataset", "twentyyear_rent.rds"))
```


### Summary of numerical variables(2022)
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyr)
library(forcats)
library(here)
library(readr)
library(dplyr)
library(gt)
data_path <- here("dataset-ignore", "twok22_data.csv")
rent_data_clean <- read_csv(data_path,show_col_types = FALSE)
data_path <- here("dataset-ignore", "twok22_data.csv")
rent_data_clean <- read_csv(data_path,show_col_types = FALSE)
summary_data <- rent_data_clean %>%
  summarise(across(c(RENTGRS, ROOMS, AGE, FTOTINC, POVERTY,INCTOT,NFAMS),
                   list(Mean = ~mean(.x),
                        Median = ~median(.x),
                        Min = ~min(.x),
                        Max = ~max(.x)),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(cols = everything(),
               names_to = "summary",
               values_to = "value") %>%
  separate(summary, into = c("variable", "statistic"), sep = "_") %>%
  pivot_wider(names_from = "statistic", values_from = "value") %>%
  select(variable, Mean, Median, Min, Max)
summary_data %>%
  gt() %>%
  tab_header(
    title = "Summary Statistics for Rental Data",
    subtitle = "Includes mean, median, minimum, and maximum values"
  ) %>%
  cols_label(
    variable = "Variable",
    Mean = "Mean",
    Median = "Median",
    Min = "Minimum",
    Max = "Maximum"
  ) %>%
  fmt_number(
    columns = c("Mean", "Median", "Min", "Max"),
    decimals = 2
  ) %>%
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(12)
  )
```

#### Speical notes to some variables we have 
1. RENTGRS
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2) 
# Histogram to show distribution of monthly gross rent
ggplot(rent_data_clean, aes(x = RENTGRS)) +
  geom_histogram(binwidth = 100, fill = "blue", color = "black") +
  labs(title = "Distribution of Monthly Gross Rent",
       x = "Monthly Gross Rent ($)",
       y = "Count") +
  theme_minimal()
```
-The histogram appears to be right-skewed, with a mode somewhere before $3000. It suggests that the mean is likely to be higher than the median, which corresponds to the stats in the above section. 
-There are a few bars past the $6000 mark, which could be considered outliers.
-There's a wide range of rents; however, the majority of the data points fall below $3000, suggesting that while there are rents across a wide range.
-According to the right skewd nature, we take a log transformation to it in later on regression model. 

2.INCTOT
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rent_data_clean, aes(x = INCTOT)) +
  geom_histogram(bins = 100, fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Histogram of Total Personal Income",
       x = "Total Personal Income",
       y = "Frequency")
```
- The histogram shows a very high peak at the lower end of the income range, suggest a right-skewed distribution. This suggests that a significant portion of the population earns within this lower income bracket. But it is what I expected because it's the most common income range in the datasets and our dataset is from census. 
- -According to the right skewd nature, we take a log transformation to it in later on regression model. 

3.FTOTINC  
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(rent_data_clean, aes(x = FTOTINC)) +
  geom_histogram(bins = 100, fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Histogram of Total Family Income",
       x = "Total Family Income",
       y = "Frequency")
```
- Simliarly as the INCTOT, the histogram shows that the distribution of total family income is right-skewed, meaning most families have an income toward the lower end of the scale, with fewer families earning higher incomes. The tail might include outliers, which are families with significantly higher incomes than the typical family and the variance and standard deviation are likely to be large.
- This variable shows to be related with INCTOT, we should test the multicollineaity issue while include it in our model. 
- Considering its nature and the possible multicollinearity issue, we plan to replace INCTOT to this variable to see if it suggests any change.

4.NFAMS
```{r echo=FALSE, message=FALSE, warning=FALSE}
nfams_counts <- rent_data_clean %>%
  group_by(NFAMS) %>%
  summarise(Count = n())

ggplot(nfams_counts, aes(x = factor(NFAMS), y = Count, fill = factor(NFAMS))) +
  geom_bar(stat = "identity") +
  labs(title = "Bar Chart of Number of Family Members",
       x = "Number of Family Members",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
The bar representing one family member is significantly taller than the others, indicating that single-person families are most common in this dataset. There is a rapid decrease in frequency as the number of family members increases. The families with a very large number of members are quite rare. This is what I expected and  aligns with my understanding to common demographic trends.

