---
title: "Class Blog 2"
subtitle: "This page provides the cleaning code for our dataset and a description of the first dataset we used. We also explore the topic we intend to investigate on this page. "
author: "Yating Zhu,Dingyu Guo,Suh Edwin Jiho"
date: "2024-03-23"
date-modified: "2024-03-23"
draft: FALSE
---

## Our Topic
Our research aims to utilize 2022 U.S. data to determine whether income influences Americans' rental choices and to explore the impact of race, age, and other factors.

## Source

The data comes from [IPUMS USA](https://usa.ipums.org/usa/index.shtml).

The original data of this website came from censuses as well as the American Community Surveys (ACS) that began in 2000 and are still going on today. We chose our dataset from this website because this website contains datasets that are large, valid and complex, containing detailed microdata. This site aims to facilitate transformative research, education, and policy-making by aggregating data across times and places, which makes the datasets on this website very appropriate for our research while streamlining our analysis of changes, data merging, and comparative research.

## Other Research on Similar Topics

### A U.S. Census Bureau reasearch: "Share of Income Needed to Pay Rent Increased the Most for Low-Income Households From 2019 to 2021"

#### From [Census.Gov](https://www.census.gov/library/stories/2023/03/low-income-renters-spent-larger-share-of-income-on-rent.html)

Research conducted by the U.S. Census Bureau from 2019 to 2021 indicates a widespread trend of increased rent burden among renter households across all income levels. Notably, the study shows that the lowest-income quantile experienced the most significant impact, with the number of cost-burdened renter-occupied households rising from 85.5% to 87.3%, equating to approximately 10.9 million households in 2021.

### A Harvard research: "More Renters Than Ever Before Are Burdened by the Rent They Pay"

#### From [NEW YORK TIMES](https://www.nytimes.com/2024/01/25/realestate/rent-prices-housing.html#)

The Harvard report also highlights this critical issue in the housing sector, revealing a substantial increase in rental burden across the United States using 2022 census data and draws similar concern as the study of U.S. Census Bureau. It found that an unprecedented 22.4 million households now allocate over 30% of their income to rent, with 12.1 million of U.S. households spending more than half of their income on this expense. This situation underscores the growing challenge of affordable housing, affecting renters' financial stability and quality of life.

These studies reveal that low-income renters face a significant rent burden, guiding the creation of targeted affordable housing initiatives, for example like rental assistance programs tailored to prevent excessive rent spending, and housing policy reforms like rent control and incentives for low-income housing development. Underpinning the vital research on rent and income relationships effectively combat the housing affordability crisis. It underscores the commitment to ensuring everyone has access to safe, affordable housing and bolsters the economic resilience of low-income families by mitigating their rent-related financial pressures.

## Reference:

Kaysen, R. (2024, January 25). More renters than ever before are burdened by the rent they pay. The New York Times. https://www.nytimes.com/2024/01/25/realestate/rent-prices-housing.html

Peter J. Mateyka and Jayne Yoo. (2023b, February 28). Share of income needed to pay rent increased the most for low-income households from 2019 to 2021. Census.gov. https://www.census.gov/library/stories/2023/03/low-income-renters-spent-larger-share-of-income-on-rent.html

## Data Loading and Cleaning

### Variables we have

| Variable     | Description                     | Notes                                                                                                   |
|---------------|--------------------|-------------------------------------|
| **RENTGRS**  | Monthly gross rent              |                                                                                                         |
| **INCTOT**   | Total personal income           | Specific codes detailed below:                                                                          |
|              |                                 | 0000001 = \$1 or break even (2000, 2005-onward ACS and PRCS)                                            |
|              |                                 | 9999999 = N/A                                                                                           |
|              |                                 | 9999998 = Unknown                                                                                       |
| **FTOTINC**  | Total family income             | Specific codes detailed below:                                                                          |
|              |                                 | -000001 = Net loss (1950)                                                                               |
|              |                                 | 9999998 = Not ascertained (1950)                                                                        |
|              |                                 | 9999999 = N/A                                                                                           |
| **HHINCOME** | Total household income          |                                                                                                         |
| **KITCHEN**  | Kitchen or cooking facilities   | 0 = N/A                                                                                                 |
| **ROOMS**    | Number of rooms                 | 00 = N/A                                                                                                |
| **BEDROOMS** | Number of bedrooms              |                                                                                                         |
| **NFAMS**    | Number of families in household |                                                                                                         |
| **AGE**      | Age                             |                                                                                                         |
| **MARST**    | Marital status                  |                                                                                                         |
| **EMPSTAT**    | Employment status             |     0=N/A                                                                                                    |
|              |                                 | 9 = Unknown/Illegible                                                                              |
| **POVERTY**  | Poverty status                  | 000 = N/A                                                                                               |


Data cleanning code:
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(here)
library(dplyr)
data_path <- here("dataset-ignore", "twok22_data.csv")
library(readr)
rent_data <- read_csv(data_path,show_col_types = FALSE)
kable(head(rent_data))
library(kableExtra)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Clean the data
library(tidyverse)
library(here)
data_path <- here("dataset-ignore", "twok22_data.csv")
library(readr)
rent_data <- read_csv(data_path,show_col_types = FALSE)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
kable(rent_data)
rent_data_clean <- rent_data %>%
  filter(RENTGRS != 0,
         !is.na(RENTGRS),
         EMPSTAT != "0" & EMPSTAT != "9",
         KITCHEN != "0", # Filter out rows where Kitchen is "0"
         ROOMS != "00", # Filter out rows where Rooms is "00"
         !INCTOT %in% c(1, 9999999, 9999998, -000001, "FTOTINC-000001"), # Filter based on INCTOT criteria
         !FTOTINC %in% c(9999998, 9999999), # Filter based on FTOTINC criteria
         POVERTY != 000
         ) # Filter based on POVERTY criteria
rent_data_clean$REGION_CLASSIFIED <- case_when(
  rent_data_clean$REGION %in% c(11, 12, 13) ~ "NORTHEAST",
  rent_data_clean$REGION %in% c(21, 22, 23) ~ "MIDWEST",
  rent_data_clean$REGION %in% c(31, 32, 33, 34) ~ "SOUTH",
  rent_data_clean$REGION %in% c(41, 42, 43) ~ "WEST",
  TRUE ~ "OTHER"
)

saveRDS(rent_data_clean, file = here("dataset", "latest.RData"))
write_csv(rent_data_clean, file = here("dataset", "latest.csv"))
```



We exclude data entries characterized by specific codes indicative of missing or unknown values, as well as those representing values deemed inappropriate for analytical purposes. Furthermore, for our dependent variable—rent—we exclude records indicating a rent of 0. This exclusion is based on the understanding that a 0 rent suggests the individual owns their home or lives in a situation not pertinent to our target research population, which is individuals who rent their housing. Given the original dataset comprises millions of samples, the exclusion of these data points still leaves us with the sufficiently cleaned large dataset for robust analysis.
