---
title: "Blog Post 7"
author: "Yating Zhu, Dingyu Guo"
date: "2024-04-19"
date-modified: "2024-04-19"
draft: FALSE
---
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyr)
library(broom)
library(dplyr)
library(here)
library(gt)
library(purrr)
data_path <- here("dataset", "data_twenty_clean.csv")
data_path1 <- here("dataset", "latest_clean.csv")
library(readr)
data_twenty <- read_csv(data_path,show_col_types = FALSE) 
data_2022 <- read_csv(data_path1,show_col_types = FALSE) 
library(ggplot2)
library(knitr)
library(lme4)
library(broom.mixed)
library(nlme)

```


```{r}
data_twenty$INCTOT <- scale(data_twenty$INCTOT, center = TRUE, scale = TRUE)
data_twenty$ROOMS <- scale(data_twenty$ROOMS, center = TRUE, scale = TRUE)
data_twenty$NFAMS <- scale(data_twenty$NFAMS, center = TRUE, scale = TRUE)
data_twenty$AGE <- scale(data_twenty$AGE, center = TRUE, scale = TRUE)
data_twenty$FTOTINC <- scale(data_twenty$FTOTINC, center = TRUE, scale = TRUE)

model_with_income <- lmer(RENTGRS ~ INCTOT + KITCHEN + ROOMS + NFAMS + AGE + MARST + RACE + EMPSTAT + FTOTINC + (1 | REGION_CLASSIFIED), data = data_twenty, REML = FALSE)
```

### Regression Coefficients Table:
```{r echo=FALSE, message=FALSE, warning=FALSE }
tidy_lm <- tidy(model_with_income)
kable(tidy_lm, format = "pipe")
```

### Model Fit Statistics Table:
```{r echo=FALSE, message=FALSE, warning=FALSE}
glance_lm <- glance(model_with_income)
kable(glance_lm, format = "pipe")
```

### Residuals and Related Statistics Table:
```{r echo=FALSE, message=FALSE, warning=FALSE}
augment_lm <- augment(model_with_income)
kable(head(augment_lm), format = "pipe")
```
- From the above tables, we observe that the term	KITCHENYes (access to kitchens) is negative in the estimate, and the variable has a large statistic, which might indicate statistic, which might indicate significance, but consider whether this variable is theoretically important for the rent price. Also AGE (Age of the renter or the rental unit): The effect size is small and the statistic value is close to zero, which may suggest it's not a strong predictor in the presence of other variables.
- Therefore we decided to remove these two variables and continue to explore the model.
### Multicollinearity concerns
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(car)
vif_values <- vif(model_with_income)
kable(vif_values)
```
- we observe income and family income stands out have higher multicollinearity index than others, therefore we decide to remove family income in our model.

```{r}
# Calculate residuals and fitted values
residuals_twenty <- resid(model_with_income)
fitted_values_twenty <- fitted(model_with_income)

# Create a residual plot
plot(fitted_values_twenty, residuals_twenty, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residual Plot", pch = 20)
abline(h = 0, col = "red") 
```




## log-log model
```{r}
# Filter out non-positive values before log transformation
data_twenty <- data_twenty[data_twenty$RENTGRS > 0 & data_twenty$INCTOT > 0, ]

# Apply log transformation
data_twenty$log_RENTGRS <- log(data_twenty$RENTGRS)
data_twenty$log_INCTOT <- log(data_twenty$INCTOT)


# Calculating means and standard deviations for the log-transformed columns
mean_log_RENTGRS <- mean(data_twenty$log_RENTGRS, na.rm = TRUE)
sd_log_RENTGRS <- sd(data_twenty$log_RENTGRS, na.rm = TRUE)
mean_log_INCTOT <- mean(data_twenty$log_INCTOT, na.rm = TRUE)
sd_log_INCTOT <- sd(data_twenty$log_INCTOT, na.rm = TRUE)


# Defining lower and upper bounds for both variables
lb_log_RENTGRS <- mean_log_RENTGRS - 3 * sd_log_RENTGRS
ub_log_RENTGRS <- mean_log_RENTGRS + 3 * sd_log_RENTGRS
lb_log_INCTOT <- mean_log_INCTOT - 3 * sd_log_INCTOT
ub_log_INCTOT <- mean_log_INCTOT + 3 * sd_log_INCTOT

# Filtering out outliers
data_twenty_clean <- data_twenty[data_twenty$log_RENTGRS >= lb_log_RENTGRS & data_twenty$log_RENTGRS <= ub_log_RENTGRS & 
                             data_twenty$log_INCTOT >= lb_log_INCTOT & data_twenty$log_INCTOT <= ub_log_INCTOT, ]
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
model_with_income_log <- lmer(log_RENTGRS ~ log_INCTOT + ROOMS + NFAMS + MARST + RACE + EMPSTAT + (1 | REGION_CLASSIFIED), data = data_twenty_clean, REML = FALSE)
```

### Regression Coefficients Table:
```{r echo=FALSE, message=FALSE, warning=FALSE }
tidy_lm <- tidy(model_with_income_log)
kable(tidy_lm, format = "pipe")
```

### Model Fit Statistics Table:
```{r echo=FALSE, message=FALSE, warning=FALSE}
glance_lm <- glance(model_with_income_log)
kable(glance_lm, format = "pipe")
```

### Residuals and Related Statistics Table:
```{r echo=FALSE, message=FALSE, warning=FALSE}
augment_lm <- augment(model_with_income_log)
kable(head(augment_lm), format = "pipe")
```



### Residual plot
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Calculate residuals and fitted values
residuals_twenty <- resid(model_with_income_log)
fitted_values_twenty <- fitted(model_with_income_log)

# Create a residual plot
plot(fitted_values_twenty, residuals_twenty, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residual Plot", pch = 20)
abline(h = 0, col = "red") 
```




